\documentclass[a4paper,12pt]{article}
\usepackage[osf]{mathpazo}
\usepackage{ms}
\usepackage{natbib}
\usepackage{lineno}
\usepackage{graphicx}
\usepackage{caption}
\modulolinenumbers[5]
\linenumbers

\pdfminorversion=3

\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

\title{Statistical and conceptual challenges in the comparative analysis of principal components}
\author{
Josef C. Uyeda$^{1,*}$, Daniel S. Caetano$^1$, and Matthew W. Pennell$^1$
}

\date{}
\affiliation{
 $^{1}$ Department of Biological Sciences \& Institute for Bioinformatics and Evolutionary Studies, University of Idaho, Moscow, ID 83844, U.S.A.\\ 
 $^{*}$ Email for correspondence: \texttt{pseudacris@gmail.com}\\
}

\runninghead{PCA in comparative analyses}
\keywords{Phylogenetic comparative methods, principal components, Brownian motion, Ornstein-Uhlenbeck, multivariate statistics}


\begin{document}

\mstitlepage
\parindent=1.5em
\addtolength{\parskip}{.3em}
\vfill

\section{Abstract}
\begin{enumerate}
\item A common procedure when analyzing multivariate data in a phylogenetic framework is to first transform the data into a set of principal component (PC) axes. Techniques for ``correcting for phylogeny'' when computing PCs have been developed and applied widely in the empirical literature. However, standard (i.e., uncorrected) principal components continue to be widely used in comprative analyses.

\item We demonstrate that failing to include phylogenetic relationships when calculating PC scores will mislead inferences in a predictable manner. Specifically, even when data is simulated under Brownian motion (BM), the first several PC scores will appear to have evolved rapidly early in a clade's history and slowed down towards the present day.

\item The primary method for calculating ``phylogenetically corrected'' PC scores relies on the assumption that the traits have evolved under a multivariate BM model. It is unknown whether this assumption is reasonable if the actual traits have evolved under more complex scenarios. Using simulations, we find that for many common models of evolution, calculating PC scores based on a BM model is a reasonable approximation though this approximation gets worse as the dimensionality of the analysis increases.

\item \emph{Synthesis:} While incorporating phylogeny when calculating PC scores does remove bias in many cases, a more fundamental issue remains when analyzing PCs in a comparative framework --- that drawing meaningful evolutionary inferences from models fit to PCs is challenging. Alternatives approaches for modeling multivariate data on a phylogeny are likely to be more informative, though further statistical and conceptual innovations will be necessary for these to be more broadly applicable.
\end{enumerate} 

\newpage

\section{Introduction}
The units of measurements are not neutral or arbitrary; different ways of representing the same set of data can change the meaning of the measurements and alter the interpretations of subsequent statistical analyses \citep{Hand2004, HansenHoule2008, Houle2011}. Therefore, careful attention needs to be paid to how data is measured and how it is transformed. Furthermore, this should be done with the scientific question in mind in order to ensure that resulting inferences are meaningful \citep{Houle2011}. 

Most approaches for studying evolution in a phylogenetic comparative framework were developed for univariate data \citep[reviewed in][]{PennellHarmon}. Even classical approaches for testing for correlated evolution across a phylogeny \citep[e.g.,][]{Felsenstein1985, Grafen1989, HarveyPagel1991} fundamentally model each trait as having evolved under an evolutionary process that is independent of the state of the other \citep{HansenOrzack2005}. However, the traits that we are actually interested in studying have likely evolved in concert with a suite of other traits and considering each trait is isolation may give a distorted picture of reality. Therefore, it is often desirable to reduce the multivariate dataset into a set of linear combinations of traits that are orthogonal to one another, such that each linear combination can then be analyzed with common comparative methods.

The most common method for reducing the dimensionality of the dataset is to perform Principal Components Analyses (PCA) prior to analyzing the data using phylogenetic comparative methods. The first PC axis is the eigenvector in the direction of greatest variance, the second PC axis, the second greatest variance, and so on. However, standard methods for calculating PC scores assume that the samples are independent of one another, which is hardly ever the case for comparative data --- as a result of shared common ancestry, relatives are likely to share many traits and trait combinations. This fact is of course, now almost universally recognized by biologists and conducting comparative analyses without considering the phylogenetic relationships of species is anathema to most evolutionary biologists. 

However, the necessity of considering phylogeny in some types of data transformations \citep{Revell2008} is not similarly recognized. Many researchers --- most of whom would never consider analyzing comparative data in a non--phylogenetic context --- continue the practice of conducting comparative analyses on standard PCs and thus inadvertantly introduce substantial bias into their analyses. Typical examples of traits where this is done include geometric morphometric landmarks \citep[e.g.,][]{Dornburg2011, Hunt2013}, measurements of multiple morphological traits \citep[e.g.,][]{Harmon2010, BergmannIrshick2012, Weir2012, Pienaar2013, Price2014}, and climatic variables \citep[e.g.,][]{KozakWiens2010, Schnitzler2012}. We stress that the papers that we have cited here are simply examples (picked more--or--less, at random) from a substantial number of papers where this non--intuitive mistake was made.

Fortunately, multiple potential solutions exists to address this problem.  One frequently used method is that of \citet{Revell2008}, who developed a procedure (explained in detail below) for converting multivariate data into ``phylogenetically corrected'' PC axes (hereafter, PPCA), by assuming that the original data evolved along the phylogeny using a multivariate Brownian motion (BM) model of evolution. In a brief simulation study, \citet{Revell2008} demonstrated that if the underlying model for the traits was indeed a multivariate BM model, performing standard PCA gave biased estimates of the eigenvalues, whereas PPCA removed this bias.

In this brief contribution we first, extend the argument of \citet{Revell2008} and demonstrate that not only are the eigenvalues obtained from PCA biased but that they are biased in a systematic and predictable way. Performing comparative analyses on standard PC axes positively misleads inference. This point has been made in other fields that deal with autocorrelation between observations, such as population genetics \citep{Novembre}, ecology \citep{Podani2002}, climatology \citep{Richman1986} and paleobiology \citep{Bookstein2012}. However, the connection between these previous results and phylogenetic comparative data has not been explicitly made and standard PCs continue to be widely used in the field. We hope that our paper helps change this practice.

Second, as stated above, \citet{Revell2008} assumed that the measured traits had evolved under a multivariate BM model. This has the potential to introduce an odd circularity into the analysis. If we assume that the underlying traits have evolved under BM and calculate a set of PPC axes based on this assumption, it seems reasonable to suppose that if the true model of trait evolution were different, the PPC scores would be inferred to have evolved in a more BM--like manner than the traits themselves. To our knowledge, the effect of this bias has not been explored. We perform a small simulation study to investigate this effect.

Last, we consider the interpretation of evolutionary models fit to PPC axes and discuss the conceptual and statistical advantages and disadvantages of using PPCA compared to alternative approaches for studying multivariate evolution in a phylogenetic comparative framework.

\section{Methods}
\subsection{\emph{Overview of PPCA}}
We briefly review the approach of \citet{Revell2008} for computing PPC scores for each species. If we assume that the traits have each evolved under a model of trait evolution, we can estimate an evolutionary variance--covariance (vcv) matrix between the $m$ traits that have been measured for $n$ species \citep{RevellHarmon2008, RevellCollar2009}:
\begin{equation}\label{eq:r}
\mathbf{R} = \frac{(\mathbf{X} - \mathbf{1a}^\prime ) \mathbf{\Sigma}^{-1} 
(\mathbf{X} - \mathbf{1a}^\prime )}{n-1}.
\end{equation}
Here $\mathbf{X}$ is the $n \times m$ matrix of observed data, $\mathbf{\Sigma}$ is a $n \times n$ matrix with elements $\Sigma_{ij}$ being the expected covariance between traits in lineages $i$ and $j$, and $\mathbf{a}$ is a vector of phylogenetic means for all $m$ traits ($\mathbf{a}=[(\mathbf{1}^\prime \mathbf{\Sigma}^{-1} \mathbf{1})^{-1} 
\mathbf{1}^\prime \mathbf{\Sigma}^{-1} \mathbf{X}]^\prime$). \emph{[include plain language description of what $\mathbf{R}$ actually represents.]}
Under a BM model of trait evolution, $\mathbf{\Sigma}$ simply mirrors the structure of the phylogeny, such that $\Sigma_{i,j}$ is the sum of shared branch lengths (from the root to the most recent common ancestor) between lineages $i$ and $j$. (We note that as only relative branch lengths matter for these calculations, we can arbitrarily set the BM rate parameter $\sigma^2$ equal to unity.) 

\emph{[Nice explanation of how eigenvalues $\mathbf{D}$ and eigenvectors $\mathbf{V}$ of the phylogenetic]}

We obtain PC scores for each species by projecting the PC axes into the original space
\begin{equation}
\mathbf{S} = (\mathbf{X} - \mathbf{1a}^\prime )\mathbf{V}
\end{equation}
and the loadings $L$ for each trait $i$ on the $j^{\mathrm{th}}$ PPC axis
\begin{equation}
L_{ij} = \frac{K_{ij}}{\sqrt{R_{ii}D_{jj}}}
\end{equation}
where $\mathbf{K}$ is the cross--covariance matrix $\mathbf{K}=(n-1)^{-1}(\mathbf{X}-\mathbf{1a}^\prime)^\prime \mathbf{\Sigma}^{-1} \mathbf{S}$ \citep[see][for details and explanation]{Revell2008}.

The end result of this procedure is a set of PPC axes, each representing of linear combinations of traits, and scores for all species for each PPC axis. The PPC scores are in the original species space and are therefore phylogenetically non--independent in the same way that the measured traits are.  


\subsection{\emph{Simulations}}

We did x

\subsection{\emph{Empirical examples}}

\section{Results}

\subsection{\emph{Simulations}}

\subsection{\emph{Empirical examples}}

\section{Discussion}
\subsection{\emph{Standard PCA should not be used for comparative analyses}} 

Discussion of main results (both simulated and empirical)

Tie into results from other studies (e.g., novembre)

These results certainly do not imply that the biological inferences made from analyzing standard PC scores in a comparative framework are necessarily incorrect.  
Interestingly, when \citet{Harmon2010} analyzed the evolution of PC2 (what they referred to as ``shape'') obtained using standard PCA, they found very little support for the ``Early Burst'' model \citep{Blomberg2003} across their 39 datasets. The fact that their use of standard PC axes biased their results \emph{towards an ``Early Burst'' pattern} only serves to strengthen their overall conclusion that such slowdowns are indeed rare \citep[but see][]{SlaterPennell}. However, our results do suggest that in some cases, analyses conducted with PC axes should perhaps be revisited to ensure that results are robust.

\subsection{\emph{Using PPCA when trait evolution is non-Brownian}}
In his original PPCA method, \citet{Revell2008} assumed that the original traits evolved under a multivariate BM model. He also suggested that one could potentially use alternative covariance structures. For example, one could first optimize the $\lambda$ model \citep{Pagel1999} across all traits simultaneously \citep[using the method of][]{Freckleton2002} and then rescale the branch lengths of the tree according to the estimated parameter in order to obtain $\mathbf{\Sigma}$ for use in Equation \ref{eq:r}. However, this is not a completely satisfying solution for two reasons. First, as noted by \citet{Revell2008}, the value of $\lambda$ used to construct the covariance structure for the PPCA will likely be different from the same parameter estimated using the PC scores themselves. Second, while this method could be extended to include other models \citep{ButlerKing2009, motmot}, this is restricted to models that assume a shared mean and variance structure across traits \citep[see][for examples where this does not  apply]{Hansen2008, Bartoszek2012}.

In our simulations, we found x.

This is because of y.

The effect of dimensionality

Note that we only simulated under a very restricted set of conditions.

\subsection{\emph{The interpretation of PPCA}}

Aside from the statistical issues raised here, a broader, and more interesting, question is how one should draw evolutionary inferences from models of trait evolution fit to PPC axes. While principal components are convenient representations of multivariate data in univariate space, this comes at the cost of abstracting the data that is analyzed from the questions that are asked. 

First, using principal components necessarily divorces model fitting from the study of important and interesting evolutionary processes. If we treat the PC scores as a univariate trait than we are only considering divergence along the PC axis. This is sensible from a purely statistical standpoint but in doing so, we may miss a great deal about what is actually going on in the data. The first principal component axis computed from comparative data is the major axis of divergence across a clade (also known as the ``line of divergence''). This axis is of considerable interest in evolutionary biology. This axis might be affected by the orientation of within--population additive genetic (co)variance $\mathbf{G}$, such that evolutionary trajectories may be biased along ``genetic lines of least resistance'' \citep[i.e., divergence occurs primarily along the leading eigenvector of $\mathbf{G}$, $G_{\text{max}}$;][]{Schluter1996, Arnoldetal2001}. Alternatively, the line of divergence may align with the ``selective lines of least resistance'', due to the structure of phenotypic adaptive landscapes \citep{Arnold2003, Jonesetal2007, Arnoldetal2008}, or else may be driven by patterns of gene flow between populations \citep{Guillaume2007} or the pleiotropic effects of new mutations \citep{Jonesetal2007, Hether2013, Houle2013}. Furthermore, pleiotropy and the genetic covariance between traits has been proposed as a potential resolution to the ``paradox of stasis'' \citep{HansenHoule2004}; traits with abundant within--population additive genetic variation may be constrained over macroevolutionary time scales \citep{Kirkpatrick2009, WalshBlows2009}. Phylogenetic comparative data can potentially provide unique and novel insights into the connection between micro-- and macroevolution \citep{Hohenlohe2008} but such questions can only be addressed in a truly multivariate context. While the traits identified by PCA will be statistically orthogonal (assuming that the model of evolution used for the transformation is appropriate; see above), this only true in the particular snapshot captured by comparative data and should not be taken to imply that they are evolving independently. 

%% Replaced this paragraph
%The first principal component axis of comparative data can be considered the major axis of diversification of a clade (often called the ``line of divergence"). This axis is of considerable interest to evolutionary biology, as it may be affected by the structure of within-population genetic variation, or the G-matrix ($\mathbf{G}$) \citep{Schluter1996}. The leading eigenvector of the additive genetic variance-covariance matrix ($G_{max}$) may bias evolution along ``genetic lines of least resistance" \citep{Schluter1996, Arnoldetal2001}. Alternatively, it is possible that alignment (or mis-alignment) between the eigenvectors of $\mathbf{G}$ and the line of divergence can result from ``selective lines of least resistance", which describe the path of least resistance along phenotypic adaptive landscapes \citep{Arnold2003, Jonesetal2007, Arnoldetal2008}. Potentially complicating the issue even further, continued gene flow and migration can result in alignment of divergent populations \citep{Guillaume2007} as can the structure of the mutational variance-covariance matrix \citep{Jonesetal2007, Hether2013}. For these reasons, testing for alignment or mis-alignment of $G_{max}$ and the line of divergence is of significant interest toward connecting micro- and macroevolution \citep{Hohenlohe2008}. However, such questions are not easily removed from a truly multivariate framework and our intuitions based on two and three trait cases do not scale well to higher dimensions. We have shown that reducing highly multivariate stochastic evolutionary models to a few principal components runs the risk of conflating evolutionary noise with evolutionary process; making interpretation of parameters and model-fits difficult at best, and abstracting the traits under study from the models being tested. 

For some macroevolutionary questions, evolutionary processes may not be of primary interest and the loss of information when using PCs may be acceptable. For instance, if the goal of fitting and comparing models is to make broad inferences about evolutionary patterns, such as the ``rate'' of trait evolution \citep{Freckleton2011, Hunt2012}, then we may not care if the models are connected to any process. Many commonly used trait models \citep[notably, the ``Pagel'' tree transformations $\lambda, \delta, \kappa$;][]{Pagel1997, Pagel1999} have no real bearing on any evolutionary processes anyways \citep{HansenOrzack2005} and therefore using PCs in conjunction with these models therefore does not seem to come at a great cost. Other models, can be tied to specific evolutionary processes \citep{HansenMartins1996, EstesArnold2007, Hansen2008, Hansen2012SysBio, PennellHarmon, PennellPE} and using PC axes necessarily breaks this tie.

Second, phenotypes are the ``stuff'' of evolution, not PC axes. Researchers often use broad terms to translate the PC axes back to trait space, such as stating that PC1 represents size and PC2, shape \citep[e.g.,][]{Harmon2010, Price2014}. However, this is a very imprecise way of quantifying biologically interesting variation. It is often difficult to understand exactly what we are explaining when we find that the evolution of PC2 is well described by, say, an OU process.

\subsection{\emph{What are the alternatives?}}

A number of alternative approaches may be used to study the evolution of correlated phenotypes along a phylogeny. The most conceptually straightforward alternative is to construct models in which there is a covariance in trait values between species (which is done in univariate models) and a covariance between different traits. Such multivariate extensions of common quantiative trait models have been developed \citep{RevellHarmon2008, Hohenlohe2008, RevellCollar2009, ButlerKing2009, motmot}. These allow researchers to investigate the connections between lines of divergence and within--population evolutionary parameters \citep{Hohenlohe2008} as well as to study how the correlation structure between traits itself changes across the phylogeny \citep{RevellCollar2009}. 

However, these approaches also have substantial drawbacks. First, the number of free parameters of the models rapidly increases as more traits are added \citep{RevellHarmon2008}, making them impractical for large mutlivariate datasets. Second, these models allow for inference of the covariance between traits but the cause of this covariance is usually not tied to specific evolutionary processes. The first issue may be addressed by constraining the model in meaningful ways \citep{ButlerKing2009} or by assuming that \emph{all traits} have the same covariance structure \citep{Adams2014, Adams2014b}; similarly, one could divide the traits into sets a priori and then assume that all traits in a set have a common variance structure \citep{Klingenberg2013}. Such restrictions of parameter space are especially appropriate for truly high--dimensional traits, such as shape inferred from geometric morphometric landmark data, for which we are primarily interested in the evolution of the aggregate trait and not necessarily the individual components. The second (inferential) difficulty can be addressed by explicilty modeling the evolution of some traits as a response to evolution or others. Hansen and colleagues have developed a number of models in which a predictor variable evolves via some process and a response variable tracks the evolution of the first as OU process \citep{Hansen2008, Labra2009, Hansen2012SysBio, Bartoszek2012}. This has been a particularly useful way of modeling the evolution of allometries \citep{Hansen2012SysBio, Voje2013, Voje2013b}. However, like the ``covariance''  models discussed above, increasing the number of traits makes the model much more complex and difficult to fit.

%For analyses of multivariate trait data, it is fairly straightforward to construct multivariate forms of commonly used trait models (e.g., BM, OU, etc.). In the univariate case, one must estimate a vcv matrix ($\mathbf{\Sigma}$ above) representing the expected variances and covariances between taxa for the trait being studied \citep[see][for details]{Harmon2010, Omeara2012}. In the multivariate case, $\mathbf{\Sigma}$ is the expected variances and covariance between all taxa and all traits \citep{RevellHarmon2008, Hohenlohe2008, RevellCollar2009, ButlerKing2009, motmot}. For the simple case of multivariate BM, fitting this model requires computing the Kronecker product of $\mathbf{R}$ (from Equation \ref{eq:r}) and $\mathbf{C}$, the patristic distance matrix derived from the phylogeny (that is, $\mathbf{\Sigma} = \mathbf{R} \otimes \mathbf{C}$). $\mathbf{\Sigma}$ is therefore a $nm \times nm$ matrix (where as above, $n$ is the number of species in the phylogeny and $m$ is the number of measured traits). While parameters may be constrained ,such as forcing several traits to share the same variance structure in an OU model, clearly the number of parameters increases very quickly as more traits are included.

%A more biologically realistic alternative is to directly model the evolution of one trait as a response to the evolution of other traits (rather than just testing whether they are correlated). Hansen and colleagues have developed a number of models in which a predictor variable evolves via some process and a response variable tracks the evolution of the first as OU process \citep{Hansen2008, Labra2009, Hansen2012SysBio, Bartoszek2012}. This has been a particular useful way of modeling the evolution of allometries \citep{Hansen2012SysBio, Voje2013, Voje2013b}. However, like the correlational models discussed above, increasing the number of traits makes the model much more complex and difficult to fit.

For such reasons, it often remains necessary to reduce the dimensionality of a multivariate dataset to one or a few compound traits. We argue that PCA can be potentially quite usefully applied to this problem, though in ways that are statistically and conceptually distinct from how it is conventionally applied to comparative data. One potential approach is to use principal components computed from within--population data, rather than comparative data. For example, if $\mathbf{G}$ (or failing that, the phenotypic variance--covariance matrix $\mathbf{P}$) is available for a focal species, then the principal axes of variation in a focal species can be measured across all species in the phylogeny. Therefore, all measurements are defined relative to a particular species' primary axes of genetic variation. This removes biases in selection of PCs that result from phylogenetic structure and stochastic evolutionary models, and focuses the study on a trait with a known relationship to other traits (at least for a focal species). 

Another alternative is to use PCA as an exploratory tool rather than as a transformation in the comparative analysis itself. For example, very often the first principal component of a multivariate morphological dataset will correlate to overall size, while the second and third will correspond to identifiable contrasts between morphological measurements. However, this correlation between PCs and applied meaning is necessarily imperfect. Thus, one could perform PCA on comparative data and use it to suggest traits that can be correlated to the major axes of variation. If the primary axis appears well--correlated to a measure of overall body size, then the researcher could design a metric that averages the scaled values of morphometric measurements, then check to see if this metric correlates well with PC1. If so, then this scaled measure of body size can be used in analyses rather than PC1 directly. Similarly, a particular axes may correspond well to a contrast between measurements of particular trait modules, for example, forelimb and hindlimb length. Thus, rather than using PCs directly, PCA can be used as a tool to select traits with known biological meaning to be used in subsequent analyses.

Of course, using either PCs defined by within--population variance structure or traits suggested by PCA will result in traits that explain less variance across taxa then PPCA. However, the extra variance explained by the principal components of comparative data may in fact include a sizeable amount of stochastic noise, rather than interesting biological trait variation (as we have shown in our simulations). While perhaps not truly orthogonal, we argue that the added intepretability of analyzing carefully--chosen (perhaps with the aid of PCA) and biologically--meaningful trait combinations far outweighs the cost of including some collinearity or explaining less than maximal variance.

\section{Concluding remarks}
In this note, we sought to clarify some statistical and conceptually issues regarding the use of principal components in comparative biology. We have shown that from a statistical standpoint, failing to consider the phylogeny when performing PCA can be positively misleading. And despite the development of methods to correct for this, in our reading of the empirical literature, we have found this to be a common oversight. A secondary aim of our paper was to provoke a discussion about how we should go about analyzing comparative data. We certainly do not have the answers but believe there are some major theoretical limitations in using PCA (phylogenetic or not) to analyze comparative data. We think that we, as a field, need to think hard about what questions we can ask with multivariate comparative data and what types of measurements will allow us to address them.

\newpage

\bibliographystyle{jecol}
\bibliography{phylopca.bib}
\end{document}
