# Simulations exploring the effects of different types of PCA on model-based phylogenetic inference.

Load packages
``` {r }
require(phytools)
require(geiger)
require(phylolm)
require(TreeSim)
require(foreach)
require(doMC)
require(MASS)
```

Read in functions
``` {r }
source("PCAfunctions.R")
```

Set the number of cores for parallel analysis
``` {r }
registerDoMC(cores=4)
```

Specify the number of tips, traits and a vector specifying which traits to fit models to.
``` {r }
ntips <- 50
ntraits <- 30
trait.seq= c(1, 2, 5, 10, 20, 30)
```

This function simulates a phylogenetic tree using a birth-death model with extinction = 0. The tree is rescaled to unit height. It then simulates a trait variance-covariance matrix by simulating eigenvalues from an exponential distribution with lambda = 1/100. These are then used to simulate under multivariate Brownian motion on the phylogeny. Note that simulating the data for a large number of tips and traits is computationally intensive (requires a ntraits x ntips covariance matrix).

``` {r }
#Skipped running and just load saved simulations
#res <- foreach(i=1:100) %dopar% simandfitPCs(ntips, ntraits, trait.seq=trait.seq)
#save(res, file=paste("sims", ntips, ntraits,".rds", sep="_"))
load(paste("sims", ntips, ntraits,".rds", sep="_"))
```

Rearrange the output to be usable for plotting
``` {r }
res.all <- lapply(1:length(res[[1]]), function(x) sapply(1:length(res), function(y) res[[y]][[x]]))
res.all <- lapply(res.all, function(x){x <- t(x); colnames(x) <- colnames(res[[1]][[1]]); x})
```

Create boxplots of Akaike weights for each transformation (raw, pca, and phylo  pca) for each of the traits in trait.seq, for each of the models (BM, OU, EB).
``` {r }
par(mfrow=c(3,3), mar=c(4, 4.1, 2.1, 0.1))
boxplot.prep <- function(x){
  tmp <- sapply(res.all, function(y) y[,x])
  colnames(tmp) <- paste("",trait.seq, sep="")
  tmp
}
bx <- lapply(1:ncol(res.all[[1]]), function(x) boxplot(boxplot.prep(x), ylab="AICw",xlab="Trait", main=colnames(res.all[[1]])[x]))
```
