# Simulations exploring the effects of different types of PCA on model-based phylogenetic inference.

Load packages
``` {r }
rm(list=ls(all=TRUE))
require(phytools)
require(geiger)
require(phylolm)
require(foreach)
require(doMC)
require(MASS)
require(nlme)
require(reshape2)
require(ggplot2)
require(plyr)
require(RColorBrewer)
require(gridExtra)
```

Read in functions
``` {r }
source("./R/analysis-helper.R")
```

Set the number of cores for parallel analysis
``` {r }
registerDoMC(cores=3)
```

Specify the number of tips, traits and a vector specifying which traits to fit models to.
``` {r }
ntips <- 50
ntraits <- 20
```

Select which of the traits to study
``` {r }
trait.seq= c(1,2,5,10,15,20)
```

This function simulates a phylogenetic tree using a birth-death model with extinction = 0. The tree is rescaled to unit height. It then simulates a trait variance-covariance matrix by simulating eigenvalues from an exponential distribution with lambda = 1/100. These are then used to simulate under multivariate Brownian motion on the phylogeny. Note that simulating the data for a large number of tips and traits is computationally intensive (requires a ntraits x ntips covariance matrix).
Skipped running and just load saved simulations
``` {r }
#simdat <- foreach(i=1:100) %dopar% sim.tree.pcs.mv(ntips, ntraits, sig2dist=rexp, lambda=1/100)
#res <- foreach(i=1:100) %dopar% fitPCs(simdat[[i]], trait.seq, models=c("BM", "OUfixedRoot", "EB"))
#res <- do.call(rbind, res)
#save(res, file=paste("./output/sims","mv", ntips, ntraits, ".rds", sep="_"))

load(paste("./output/sims","mv", ntips, ntraits, ".rds", sep="_"))

bm.cor <- build.sim.data.table(res, trait.set=trait.seq)
```

Figure 1- Distribution of Akaike weights for the Brownian motion (top row), Ornstein–Uhlenbeck (middle row) or Early–Burst model (bottom row).
Data simulated under correlated multivariate Brownian Motion
``` {r }
fig.box.aicw(bm.cor)
```

# Simulate uncorrelated data under BM, OU and EB and analyze using model selection, node-height test, and disparity through time.

Simulate the data
``` {r }
nsims = 100
#bmdat <- lapply(1:nsims, function(x) sim.tree.pcs.ind(ntips, ntraits, sig2dist=function(x){0.25}))
#oudat <- lapply(1:nsims, function(x) sim.tree.pcs.ind.ou(ntips, ntraits, alpha=2, sig2=1))
#ebdat <- lapply(1:nsims, function(x) sim.tree.pcs.ind.eb(ntips, ntraits, a=log(0.02), sig2=1))
```

Calculate the contrasts for each of the first 50 simulated datasets.
``` {r }
#bmcont <- get.contrasts(bmdat[1:50], "BM")
#oucont <- get.contrasts(oudat[1:50], "OU")
#ebcont <- get.contrasts(ebdat[1:50], "EB")
```

Combine all and save
``` {r }
#contrasts <- rbind(bmcont, oucont, ebcont)
#save(contrasts, file="./output/contrastsdata.rds")
```

Calculate the disparity through time for each simulated dataset
``` {r }
#bmdtt <- get.dtt(bmdat[1:50], "BM") 
#oudtt <- get.dtt(oudat[1:50], "OU")
#ebdtt <- get.dtt(ebdat[1:50], "EB")

#dispdat <- do.call(rbind, list(bmdtt, oudtt, ebdtt))
#save(dispdat, file="./output/disparitydata.rds")
```

Load back in saved output for contrasts and disparity
``` {r }
load("./output/contrastsdata.rds")
load("./output/disparitydata.rds")
```

Figure 3- Relationship between the average phylogenetic independent contrasts and the height of the node across 100 datasets simulated under either
``` {r }
##a BM (top row), OU (middle row) or EB (bottom row) model of evolution.
fig.nh.3panel(contrasts)
```

Figure 4- Relative disparity through time for the same simulated datasets as in Figure 3.
``` {r }
fig.dtt.3panel(dispdat)
```

Simulate under OU, BM and EB uncorrelated, fit models and calculate AICweights and compare parameter estimates.
``` {r }
registerDoMC(cores=3)
#oufits <- foreach(i=1:100) %dopar% fitPCOU(oudat[[i]], 1:20, models=c("BM", "OUfixedRoot", "EB"))
#ebfits <- foreach(i=1:100) %dopar% fitPCs(ebdat[[i]], trait.seq, models=c("BM", "OUfixedRoot", "EB"))
#bmfits <- foreach(i=1:100) %dopar% fitPCs(bmdat[[i]], trait.seq, models=c("BM", "OUfixedRoot", "EB"))
```

Get just the aicws for the OU results, filtering out the parameter estimates
``` {r }
#ouaicws <- lapply(oufits, function(x) x$aicw.table)
#oures <- do.call(rbind, ouaicws)
```
Collect other AIC weights for EB and BM
``` {r }
#ebres <- do.call(rbind, ebfits)
#bmres <- do.call(rbind, bmfits)
```

Save output
``` {r }
#save(oures, file=paste("./output/simsOU","ind", ntips, ntraits, ".rds", sep="_"))
#save(ebres, file=paste("./output/simsEB","ind", ntips, ntraits, ".rds", sep="_"))
#save(bmres, file=paste("./output/simsBM","ind", ntips, ntraits, ".rds", sep="_"))
```

Load fits back into the workspace
``` {r }
load(paste("./output/simsOU","ind", ntips, ntraits, ".rds", sep="_"))
load(paste("./output/simsBM","ind", ntips, ntraits, ".rds", sep="_"))
load(paste("./output/simsEB","ind", ntips, ntraits, ".rds", sep="_"))
```

Get the parameter estimates from the OU simulations, as well as the mean AIC weights.
``` {r }
#parsdf <- get.parsOU(oufits)
#OUmeanAICw <- get.meanAICws(oufits)

#save(OUmeanAICw, file="./output/OUmeanAICw.rds")
#save(parsdf, file="./output/OUsimParameterEstimates.rds")

load("./output/OUmeanAICw.rds")
load("./output/OUsimParameterEstimates.rds")
```

Multivariate BM
``` {r }
bm.ind <- build.sim.data.table(bmres, trait.seq)
```

Boxplots for AIC weights for models for uncorrelated mvBM
``` {r }
fig.box.aicw(bm.ind)
```

Boxplots for AIC weights for models for uncorrelated mvOU
``` {r }
ou.ind <- build.sim.data.table(oures, trait.seq)
fig.box.aicw(ou.ind)
```

Boxplots for AIC weights for models for uncorrelated mvEB
``` {r }
eb.ind <- build.sim.data.table(ebres, trait.seq)
fig.box.aicw(eb.ind)
```

do a bit of processing
``` {r }
parsdf$trait <- as.character(parsdf$trait)
parsdf$trait <- sapply(parsdf$trait, function(x) sub(".", "", x, fixed=TRUE))
parsdf$trait <- factor(parsdf$trait)
pars.tmp <- melt(parsdf)
colnames(pars.tmp) <- colnames(OUmeanAICw)
```

combine the dataframes
``` {r }
est <- rep("AICw", nrow(OUmeanAICw))
OUmeanAICw <- cbind(OUmeanAICw, est)
est <- rep("Par", nrow(pars.tmp))
pars.tmp <- cbind(pars.tmp, est)
oudf <- rbind(OUmeanAICw, pars.tmp)
```

prune down dataset to raw and phylo pca only
``` {r }
oudf <- subset(oudf, type %in% c("raw", "ppc"))
oudf$type <- factor(oudf$type)
oudf$type <- factor(oudf$type, levels=c("raw", "ppc"), labels=c("Original data", "Phylogenetic PCA"))
```

prune down dataset to exclude only BM, OU, EB, alpha
``` {r }
oudf <- subset(oudf, simmodel %in% c("BM", "EB", "OUfixedRoot", "alpha"))
oudf$simmodel <- factor(oudf$simmodel)
oudf$simmodel <- factor(oudf$simmodel, levels=c("BM", "OUfixedRoot", "EB", "alpha"),
                        labels=c("BM", "OU", "EB", "alpha"))

oudf$trait <- factor(oudf$trait, levels=unique(oudf$trait), labels=c(1:20))
```


Figure showing detailed model fit results and parameter estimates when simulated under uncorrelated mvOU and fit to each of the different models.
``` {r }
fig.model.support.alpha(oudf)
```

## Matrix Rank Simulations
``` {r }
nsims = 25
ntraits = 20

# This sequence gives a reasonably smooth visualization across sets of eigenvalues where the proportion of variance explained by PC1 varies evenly between 1/20 and 1.
seqa <- c(-5, -1.5, -1, -0.75, -.5, -0.2, 0, 0.15, seq(0.3,2.1, 0.1), 2.25, 2.5, 5)
rank.seq <- exp(seqa)
#plot(seqa, sapply(rank.seq, function(x) 1/sum(ev^x)), xlim=c(-5, 5), ylim=c(0, 1))
#abline(h=sapply(rank.seq, function(x) 1/sum(ev^x)))

varEV1 <- sapply(1:length(rank.seq), function(x) 1/(sum(ev^rank.seq[x])))
#rankdat <- lapply(rank.seq, function(y) lapply(1:nsims, function(x) sim.tree.pcs.mv(ntips, ntraits, sig2dist=ev.rank, p=y)))
#rankcont <- lapply(1:length(rank.seq), function(x) get.contrasts(rankdat[[x]], x))
#rankcont <- do.call(rbind, rankcont)
#rankslopes <- ddply(rankcont, .(type, rep, simmodel, variable), summarize, slope=lmslope(value, times))
#save(rankslopes, file="./output/rankslopes.rds")

load("./output/rankslopes.rds")

fig.rankslopes(rankslopes)
```

# Empirical analyses of morphological datasets

## Analysis of Felidae dataset.

Reference for the Felids phylogeny, trees available in on-line supplementary file:
Nyakatura, K., and O. R. Bininda-Emonds. 2012. Updating the evolutionary history of Carnivora (Mammalia): A new species-level supertree complete with divergence time estimates. BMC biology 10:12.

Get only the first dated phylogeny in the file. Tree has best estimate for branch lengths.
``` {r }
phy <- read.nexus("./datasets/1741-7007-10-12-s5.nex")[[1]]
```

Morphological data manually compiled from:
Slater, G. J., and B. Van Valkenburgh. 2009. Allometry and performance: The evolution of skull form and function in felids. Journal of Evolutionary Biology 22:2278–2287.
Sakamoto, M., G. T. Lloyd, and M. J. Benton. 2010. Phylogenetically structured variance in felid bite force: The role of phylogeny in the evolution of biting performance. Journal of Evolutionary Biology 23:463–478.

Get dataset:
``` {r }
data <- read.csv("./datasets/felidae_data.csv", header = TRUE, sep = "\t")
rownames(data) <- data[,1]
```

Drop from the tree the species we do not have data for:
``` {r }
phy.fel <- drop.tip(phy, tip = phy$tip.label[which(!phy$tip.label %in% data[,1])])
phy.fel$edge.length <- phy.fel$edge.length/max(branching.times(phy.fel))
```

Model selection analysis:
``` {r }
mm <- match(phy.fel$tip.label, data[,1])
dt <- log(data[mm,2:8])
felidPC <- princomp(dt, cor=TRUE)
felidPPC <- phyl.pca(phy.fel, dt, mode="corr")

felidae.dat <- list(tree=phy.fel, raw=dt, pc=felidPC$scores, ppc= felidPPC$S,
                    pcall=felidPC, ppcall=felidPPC)
felidaeFit <- fitPCs(felidae.dat, 1:ncol(dt), models=c("BM", "OUfixedRoot", "EB"))

felidae.ind <- build.sim.data.table(felidaeFit, 1:7)
```

Felid dataset is highly correlated. PC1 explains 96.9% of the variance under PCA 93.7 under pPCA.
``` {r }
round((felidPC$sdev^2)[1] / sum(felidPC$sdev^2), digits=3)
round(diag(felidPPC$Eval)[1] / sum(diag(felidPPC$Eval)),  digits=3)
```

Model support across datasets showing AIC weights:
``` {r }
## pdf("felidae.aicw.pdf", width=9, height=7)
fig.emp.aicw(felidae.ind)
## dev.off()
```

Analysis of contrasts and disparity:
``` {r }
nsims=1; ntraits=ncol(dt)
felidCont <- get.contrasts(list(felidae.dat), "contrasts")
felidDisp <- get.dtt(list(felidae.dat), "disparity")
```

Node height and disparity through time plots for the felid dataset.
``` {r }
## pdf("felidae.nh-dtt.pdf", width=9,height=7)
emp.graph(felidCont, felidDisp, "Purples")
## dev.off()
```

# Analisis of dataset for Cyprinodon fishes.

Read the dataset and phylogeny.
``` {r }
tr <- read.nexus("./datasets/cypri_underscore.nex")
dt <- read.csv("./datasets/cypri_dataset.csv")

dt$species <- gsub(" ", "_", dt$species)
tr$tip.label <- gsub("'", "", tr$tip.label)
```

Keep only the data for Cyprinodon species.
``` {r }
sp.split <- sapply(dt$species, FUN = function(x) strsplit(x, split = "_"))
cyp <- sapply(1:length(sp.split), FUN = function(x) sp.split[[x]][1])
id <- which(cyp == "Cyprinodon")
dt <- dt[id,]
```

Some species names need correction.
``` {r }
dt$species[which(dt$species == "Cyprinodon_sp._'durophage'_San_Salvador_Island")] <- "Cyprinodon_bozo"
dt$species[which(dt$species == "Cyprinodon_sp._'scale-eater'_San_Salvador_Island")] <- "Cyprinodon_bulldog"
dt$species[which(dt$species == "Cyprinodon_sp._'detritivore'_San_Salvador_Island")] <- "Cyprinodon_normal_Crescent_Pond"
```

Check if all species in the dataset are in the phylogeny.
``` {r }
index <- which(tr$tip.label %in% dt$species)
sum(dt$species %in% tr$tip.label[index]) == dim(dt)[1]
```

Keep only the species in the dataset in the tree (Drop the non-Cyprinodon).
```{r}
tr <- drop.tip(phy = tr, tip = tr$tip.label[-index])
```

Match dataset and tree. Keep only the morphological measurements.
Data is already size corrected.
```{r}
mm <- match(tr$tip.label, dt$species)
dt <- dt[mm,-c(17:25)]
rownames(dt) <- dt$species
dt <- dt[,-1]
```

Repeat analysis done with the Felid:

Model selection:
``` {r }
cypriPC <- princomp(dt, cor=TRUE)
cypriPPC <- phyl.pca(tr, dt, mode="corr")

cypri.dat <- list(tree=tr, raw=dt, pc=cypriPC$scores, ppc=cypriPPC$S,
                    pcall=cypriPC, ppcall=cypriPPC)
cypriFit <- fitPCs(cypri.dat, 1:ncol(dt), models=c("BM", "OUfixedRoot", "EB"))

cypri.ind <- build.sim.data.table(cypriFit, 1:15)
```

Show correlation among traits. PC1 explains 37.8% and pPC1 32.3%.
``` {r }
round((cypriPC$sdev^2)[1] / sum(cypriPC$sdev^2), digits=3)
round(diag(cypriPPC$Eval)[1] / sum(diag(cypriPPC$Eval)),  digits=3)
```

Last 4 PCA and pPCA axes:
``` {r }
sum( ( (cypriPC$sdev^2) / sum(cypriPC$sdev^2) )[12:15] )
sum( ( diag(cypriPPC$Eval) / sum(diag(cypriPPC$Eval)) )[12:15] )
```

Model support across datasets showing AIC weights:
``` {r }
pdf("cypri.aicw.pdf", width=9,height=7)
fig.emp.aicw(cypri.ind)
dev.off()
```

Analysis of contrasts and disparity:
``` {r }
nsims=1; ntraits=ncol(dt)
cypriCont <- get.contrasts(list(cypri.dat), "contrasts")
cypriDisp <- get.dtt(list(cypri.dat), "disparity")
```

Node height and disparity through time plots.
``` {r }
pdf("cypri.nh-dtt.pdf", width=9,height=7)
emp.graph(cypriCont, cypriDisp, "Purples")
dev.off()
```
